### 2016-11-28 local-capability-machine.tex
Updated f2 correctness lemma
small corrections

%Variables
\newcommand{\var}[1]{\mathit{#1}}
\newcommand{\hs}{\var{ms}}
\newcommand{\ms}{\hs}
\newcommand{\hv}{\var{hv}}
\newcommand{\rv}{\var{rv}}
\newcommand{\lv}{\var{lv}}
\newcommand{\gl}{\var{g}}
\newcommand{\pc}{\mathit{pc}}
\newcommand{\pcreg}{\mathrm{pc}}
\newcommand{\addr}{\var{a}}
\newcommand{\offset}{\var{offset}}
\newcommand{\word}{\var{w}}
\newcommand{\start}{\var{base}}
\newcommand{\addrend}{\var{end}}
\newcommand{\pwlv}{\var{pwl}}
\newcommand{\mem}{\var{mem}}
\newcommand{\reg}{\var{reg}}
\newcommand{\heapseg}{\var{ms}}
\newcommand{\heap}{\var{mem}}
\newcommand{\mode}{\var{mode}}
\newcommand{\perm}{\var{perm}}
\newcommand{\permp}{\var{permPair}}
\newcommand{\roll}{\var{roll}}
\newcommand{\instr}{\var{instr}}
\newcommand{\stdcap}[1][(\perm,\gl)]{\left(#1,\start,\addrend,\addr \right)}
\newcommand{\adv}{\var{adv}}
\newcommand{\msframe}{ms_\var{frame}}
\newcommand{\link}{\var{link}}
\newcommand{\stk}{\var{stk}}
\newcommand{\flag}{\var{flag}}
\newcommand{\nwl}{\var{nwl}}
\newcommand{\pwl}{\var{pwl}}

@@

%Domains
\newcommand{\plaindom}[1]{\mathrm{#1}}
\newcommand{\Caps}{\plaindom{Cap}}
\newcommand{\Words}{\plaindom{Word}}
\newcommand{\Addrs}{\plaindom{Addr}}
\newcommand{\ExecConfs}{\plaindom{ExecConf}}
\newcommand{\RegName}{\plaindom{RegisterName}}
\newcommand{\Regs}{\plaindom{Reg}}
\newcommand{\Heaps}{\plaindom{Mem}}
\newcommand{\HeapSegments}{\plaindom{MemSegment}}
\newcommand{\Confs}{\plaindom{Conf}}
\newcommand{\Instrs}{\plaindom{Instructions}}
\newcommand{\nats}{\mathbb{N}}
\newcommand{\ints}{\mathbb{Z}}
\newcommand{\Perms}{\plaindom{Perm}}
\newcommand{\Globals}{\plaindom{Global}}

@@

\begin{document}
\begin{flushright}
\today
\end{flushright}
\section{Domains and Notation}
\begin{align*}
\Addrs &::= \nats\\
\Words &::= \Caps + \ints \\
\Regs  &::= \RegName \rightarrow \Words\\
\Heaps &::= \Addrs \rightarrow \Words \\
\Perms &::= \{ \noperm, \readonly, \readwrite, \readwritel, \exec, \entry, \rwx, \rwlx\}\\
\ExecConfs  &::= \Regs \times \Heaps \\
\Caps  &::= (\Perms \times \Globals) \times \Addrs \times (\Addrs + \{ \infty \}) \times \Addrs\\
\Confs &::= \ExecConfs + \{\failed, \halted \times \ExecConfs\} \\
\HeapSegments &::= \Addrs \parfun \Words\\
\Globals & ::= \{\glob, \local \}
\end{align*}
Local capabilities have been added by adding a new domain $\Globals$ which represents whether a capability is local or global. Two new permissions $\readwritel$ and $\rwlx$ that permits writing local capabilities. They are otherwise the same as their non-''permit write local'' counterparts.

The $\glob$ and $\local$ are ordered as shown in Figure~\ref{fig:glob-hier}.
\begin{figure}[!h]
  \centering
  \begin{tikzpicture}[main node/.style={}]
  \node[main node] (1) {$\glob$};
  \node[main node] (2) [below of=1] {$\local$};

@@

\begin{table}[!h]
  \centering
  \begin{tabular}[!h]{r |  p{7cm} }
  $\noperm$ & No permissions. Grants no permissions\\
\hline
  $\readonly$ & Read only. Grants read permission \\
\hline
  $\readwrite$ & Read-write. Grants read and write permissions. Storage of local capabilities prohibited. \\
\hline
  $\readwritel$ & read-write, permit write local. Grants read and write permissions. Storage of local capabilities granted. \\
\hline
  $\exec$ & Execute permission. Grants execute and read permissions.\\
\hline
  $\entry$ & Entry permission. This permission grants no access, but when jumped to, it Will turn into an $\exec$ permission.\\
\hline
  $\rwx$ & Read, write, execute permission. Grants read, write, and execute permissions. Storage of local capabilities prohibited. \\
\hline
  $\rwlx$ & Read, write, execute permission with permit write local. Grants read, write, and execute permissions. Storage of local capabilities possible.
\end{tabular}

@@

\caption{Permission hierarchy}
\label{fig:perm-hier}
\end{figure}
Notation:
\[
\begin{array}{rcl}
i       &\in& \Instrs \\
r       &\in& \RegName\\
\pc     &\in& \Caps \\
\pcreg  &\in& \RegName \\
\Phi    &\in& \ExecConfs \\
\memheap&\in& \Heaps \\
\memreg &\in& \Regs \\
\addr   &\in& \Addrs\\
\perm   &\in& \Perms\\
((\perm,\gl),\start,\addrend,\addr) &\in& \Caps \\
n       &\in& \ints\\
\end{array}
\]
Further definitions:
\[
\begin{array}{rcl}
\lv    &::=& \refreg{r} \\
\hv    &::=& \refheap{r}\\
\rv    &::=& n \mid \lv \\
i      &::=& 
             \jmp{\lv} \mid 
             \jnz{\lv}{\rv} \mid
             \move{\lv}{\rv} \mid 
             \load{\lv}{\hv} \mid 
             \store{\hv}{\rv} \mid  \\
       &   & \plus{\lv}{\rv}{\rv} \mid 
             \lea{\lv}{\rv} \mid 
             \restricttwo{\lv}{\rv} \mid 
             \subseg{\lv}{\rv}{\rv} \mid  \\
       &   & \isptr{\lv}{\rv} \mid 
             \getp{\lv}{\lv} \mid 
             \getl{\lv}{\lv} \mid 
             \getb{\lv}{\lv} \mid
             \gete{\lv}{\lv} \mid
             \geta{\lv}{\lv} \mid \\
       &   & \fail \mid
             \halt \mid 
\end{array}
\]
Further define $\reg_0 \in \Regs$ such that
\[
  \forall r \in \RegName \ldotp \reg_0(r) = 0
\]

\subsection*{Semantics}
Assume a $\decode$ function that decodes integer to instructions:
\begin{align*}
\decode &:\Words \fun \Instrs
\end{align*}
Assume an $\encodePerm$, $\encodeLoc$, and $\encodePermPair$ function that encodes a permissions, locality, and permission pair, respectively, as an integer:
\begin{align*}
\encodePerm &: \Perms \fun \ints \\
\encodeLoc &: \Globals \fun \ints \\
\encodePermPair &: (\Perms \times \Globals) \fun \ints \\
\end{align*}
Further, assume an inverse function, $\decodePermPair{}$, that decodes permissions
\[
  \decodePermPair{} : \ints \fun (\Perms \times \Globals)
\]
\begin{align*}
  \Phi & \rightarrow \sem{\decode(\memheap(\addr))}(\Phi) & &                                   
                                                              \arraycolsep=0pt
                                                              \begin{array}{l}
                                                                \text{if $\memreg(\pcreg) = \stdcap$}\\
                                                                \quad\text{and $\start \leq \addr \leq \addrend$}\\
                                                                \quad\text{and $\perm \in \{ \exec,\rwx, \rwlx \}$ }
                                                              \end{array}\\
\Phi & \rightarrow \failed                                 & & \text{otherwise}
\end{align*}

@@

                                                   \quad\text{and $\var{newPermPair} = \decodePermPair(n)$}\\
                                                   \quad\text{and $\var{newPermPair} \sqsubseteq \permp$}\\
                                                   \quad\text{and $c = (\var{newPermPair},\start,\addrend,\addr)$}

@@

@@

                                                \text{if $\memreg{r} = \stdcap[\permp]$} \\

@@

                                                \quad\text{and $c = (\permp,n_2,n_2,\addr)$}

@@

                              \text{if $\memreg{r_2} = \stdcap$}

@@

                              \text{if $\memreg{r_2} = \stdcap$}

@@

                              \text{if $\memreg{r_2} = \stdcap$}

@@

                              \text{if $\memreg{r_2} = \stdcap$}

@@

                              \text{if $\memreg{r_2} = \stdcap$}

@@

                            \stdUpdatePc{\updateReg{r_1}{1}} & \text{if $\memreg{\rv} = \stdcap$} \\

@@

After adding local capabilities, most of the definitions have changed slightly to take into the account new form the capabilities take. The only two instructions that have undergone enough change to deserve to be mentioned are $\store{}$ and $\restrict{}{}{}$.

$\store{}$ has the same conditions as before, but if what it tries to write is a capability, then it needs to be a write access that permits write local.

$\restrict{}{}{}$ looks like before, but now it uses a ``permission pair'' which is the access permission and whether it is local or not. Whether a $\restrict{}{}{}$ succeeds depends on the ordering, so the real change lies in the new ordering, which is described further above.

Define the following simple macros:

\begin{align*}
  \restrict{r_1}{r_2}{r_3} \; r_4 &\defeq
                                    \begin{aligned}[t]
                                      & \move{r_1}{r_2} \\
                                      & \restrict{r_1}{r_3}{r_4}
                                    \end{aligned} \\
  \subseg{r_1}{r_2}{r_3} \; r_4   &\defeq
                                    \begin{aligned}[t]
                                      & \move{r_1}{r_2} \\
                                      & \subseg{r_1}{r_3}{r_4}
                                    \end{aligned}\\
  \lea{r_1}{r_2} \; r_3           &\defeq
                                    \begin{aligned}[t]
                                      & \move{r_1}{r_2} \\
                                      & \lea{r_1}{r_3}
                                    \end{aligned}\\
  \store{r}{n} &\defeq
                   \begin{aligned}[t]
                     & \move{r_i}{n} \qquad \text{ where $r_i$ unused register}\\
                     & \store{r}{r_i}
                   \end{aligned}
\end{align*}

@@

\begin{specification}[Malloc v.2]
  $c_\malloc$ satisfies the specification for malloc iff
\[  
  \begin{aligned}
    & \exists \iota_{\malloc,0} \ldotp \\
    & \quad \forall \Phi \in \ExecConfs \ldotp \forall \ms_{\var{footprint}}, \hsframe \in \HeapSegments \ldotp \\
    &\qquad \forall i, n, \size \in \nats \ldotp \forall b,e,a \in \Addrs\ldotp\forall p \in \Perms \ldotp \\
    &\qquad \quad \forall \iota_\malloc \futurestr \iota_{\malloc,0} \land \\
    &\qquad \quad \memheap = \ms_{\var{footprint}} \uplus \hsframe \land\\
    &\qquad \quad \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
    &\qquad \quad \memreg(r_1) = \size \land \\
    &\qquad \quad \size \geq 0 \land \\
    &\qquad \quad \memreg(r_0) = (p,b,e,a) \land \\
    &\qquad \quad \memreg(\pcreg) = c_\malloc \\ %and this needs to change with ABI.
    &\qquad \quad \Rightarrow \\
    &\qquad \qquad\exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
    &\qquad \qquad \quad \exists j,k \in \nats \ldotp \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
    &\qquad \qquad \qquad \Phi \step[j] \Phi' \land \\
    &\qquad \qquad \qquad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
    &\qquad \qquad \qquad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
    &\qquad \qquad \qquad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
    &\qquad \qquad \qquad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
    &\qquad \qquad \qquad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{(p,b,e,a)}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
    &\qquad \qquad \qquad \size - 1 = e'-b' 
  \end{aligned}
\]
where $\iota_{\malloc,0}$ is the region that initially governs malloc 
\[
\forall \iota' \futurestr \iota_{\malloc,0} \ldotp \forall W,i \ldotp W(i)=\iota' \Rightarrow \iota'.H (\iota'.s) (W) = \iota'.H (\iota'.s ([i \mapsto W(i)]))
\]
and
\[
  \iota_{\malloc,0}.v = \perma
\]

\end{specification}
In the specification above $\iota_\malloc'$ is a future region of the initial region that governs malloc.
\lau{should it be a public or private future world in the premise?}

\section{Macros}
In order to write readable example programs, we provide macros that can be programmed using the instruction set given in the formalisation.
\dominique{how is the return capability constructed: not on the stack, I presume, so  means mallocing a block of memory for the activation record?}
\lau{See the new addition below - when we don't have a stack, then we have no need to ``save'' local capabilities which means that we can have the return code as part of the program. The only thing we have to do is to have the capability point to the return address and make the capability into an entry capability. }
\lau{On further thought, conveniently the examples did not seem to need an activation record, but if we want \texttt{call} to work more generally speaking, then it needs to setup an activation record. I have changed the description of \texttt{call}, so it now uses an activation record. }
\dominique{Similarly: how is malloc invoked?  Do we trust malloc enough to not encapsulate ourselves from it, i.e. provide an rx return capability and use callee-save registers?}
\lau{ I think this is a conceptual question as we can make it work with either. We already trust malloc to give us a fresh piece of memory and not reuse it later on, so we already assume malloc to be somewhat trusted, so why not go all the way? }
\dominique{what does ``fetch the capability ...'' mean?}
\lau{ We don't know where the capability resides, but if it is in memory, then it will be loaded into a register. }
\dominique{Wouldn't it be more clear to provide call with two explicit lists of registers: those which need to be stored, and those which are provided as arguments (i.e. which do not need to be erased)?}
\dominique{Perhaps you could also provide an explicit syntax for ``undefined symbols'' that should be filled in by a linker?}

\subsection{Linking and ABI}
In order to make capabilities to trusted code (and possibly untrusted code) available, we assume that some sort of linker has made these available. This is done in the following way: For every function, the first memory cell the capability for that function governs contains a capability for the list in question. Each function name in a program corresponds to an offset in the table, e.g., \texttt{malloc}. When a name is used in a program, it indicates what entry to pick from the list. The list should always be accessible by taking a copy of the capability in the $\pcreg$-register and adjusting it to point to the first cell it governs.

The capability link list can be shared between multiple functions as it is accessed through a read-only capability.

\subsection{Flag list}
A function may use flags to signal failure. We use the convention that a flag list is available in the second memory cell of a functions code (so just after the linking table). The flag list is accessed through a read-write capability and initially it contains all zero. Like the linking list, each entry is associated with a name which may appear in the macros.

The flag list should never be shared between distrusting parties.

\begin{description}
\item[\texttt{call} $f(\bar{r}_{\var{args}},\bar{r}_{\var{priv}})$] $\bar{r}_{\var{args}}$ and $\bar{r}_{\var{priv}}$ are lists of registers. An overview of this call:
  \begin{itemize}
  \item Set up activation record
  \item Create local enter capability for activation
  \item Clear unused registers
  \item Jump
  \item Upon return: Run activation code
  \end{itemize}
A more detailed description of each of the above steps:
\begin{description}
  \item [Set up activation record]
    \begin{itemize}
    \item Run malloc to get a piece of memory with space for:
    \begin{itemize}
      \item Words in $\bar{r}_{\var{priv}}$
      \item Code return capability (opc)
      \item Activation code
    \end{itemize}
    \item Store the words in $\bar{r}_{\var{priv}}$ to the activation record. 
    \item Adjust a copy of the current pc to point to the return address and save it to the activation record.
    \item Write the activation code to the activation record.
  \end{itemize}
  \item [Create local enter capability for activation] Make the capability for the activation record point to the beginning of the activation record and restrict it to a local-enter capability. Place this capability in $r_0$.
  \item [Clear unused registers]
    Clear all the register that are not $\pcreg$, $r_0$ or in $\bar{r}_{\var{args}}$.
  \item [Jump] Load the entry $f$ from the linking table to a free register and jump to it.
  \item [Activation code] The activation code does the following:
    \begin{itemize}
    \item Move the stored ``private'' words in to their respective $\bar{r}_{\var{priv}}$ registers.
%    \item Clear all registers but $r_1$, $\pcreg$, and $\bar{r}_{\var{priv}}$.
    \item Load the return address to $\pcreg$
    \end{itemize}
\end{description}
\item[\texttt{malloc $r$ $n$}] Calls malloc to allocates a piece of memory of size $n$. The capability will be stored in register \texttt{r}.

>>>>

%Variables
\newcommand{\var}[1]{\mathit{#1}}
\newcommand{\hs}{\var{ms}}
\newcommand{\ms}{\hs}
\newcommand{\hv}{\var{hv}}
\newcommand{\rv}{\var{rv}}
\newcommand{\lv}{\var{lv}}
\newcommand{\gl}{\var{g}}
\newcommand{\pc}{\mathit{pc}}
\newcommand{\pcreg}{\mathrm{pc}}
\newcommand{\addr}{\var{a}}
\newcommand{\offset}{\var{offset}}
\newcommand{\word}{\var{w}}
\newcommand{\start}{\var{base}}
\newcommand{\addrend}{\var{end}}
\newcommand{\pwlv}{\var{pwl}}
\newcommand{\mem}{\var{mem}}
\newcommand{\reg}{\var{reg}}
\newcommand{\heapseg}{\var{ms}}
\newcommand{\heap}{\var{mem}}
\newcommand{\mode}{\var{mode}}
\newcommand{\perm}{\var{perm}}
\newcommand{\permp}{\var{permPair}}
\newcommand{\roll}{\var{roll}}
\newcommand{\instr}{\var{instr}}
\newcommand{\stdcap}[1][(\perm,\gl)]{\left(#1,\start,\addrend,\addr \right)}
\newcommand{\adv}{\var{adv}}
\newcommand{\msframe}{ms_\var{frame}}
\newcommand{\link}{\var{link}}
\newcommand{\stk}{\var{stk}}
\newcommand{\flag}{\var{flag}}
\newcommand{\nwl}{\var{nwl}}
\newcommand{\pwl}{\var{pwl}}
\newcommand{\sta}{\var{sta}}
\newcommand{\cnst}{\var{cnst}}

@@

%Domains
\newcommand{\plaindom}[1]{\mathrm{#1}}
\newcommand{\Caps}{\plaindom{Cap}}
\newcommand{\Words}{\plaindom{Word}}
\newcommand{\Addrs}{\plaindom{Addr}}
\newcommand{\ExecConfs}{\plaindom{ExecConf}}
\newcommand{\RegName}{\plaindom{RegisterName}}
\newcommand{\Regs}{\plaindom{Reg}}
\newcommand{\Heaps}{\plaindom{Mem}}
\newcommand{\Mems}{\Heaps}
\newcommand{\HeapSegments}{\plaindom{MemSegment}}
\newcommand{\MemSegments}{\HeapSegments}
\newcommand{\Confs}{\plaindom{Conf}}
\newcommand{\Instrs}{\plaindom{Instructions}}
\newcommand{\nats}{\mathbb{N}}
\newcommand{\ints}{\mathbb{Z}}
\newcommand{\Perms}{\plaindom{Perm}}
\newcommand{\Globals}{\plaindom{Global}}

@@

\begin{document}
\begin{flushright}
\today
\end{flushright}
\section{Domains and Notation}
\begin{align*}
\Addrs &::= \nats\\
\Words &::= \Caps + \ints \\
\Regs  &::= \RegName \rightarrow \Words\\
\Heaps &::= \Addrs \rightarrow \Words \\
\Perms &::= \{ \noperm, \readonly, \readwrite, \readwritel, \exec, \entry, \rwx, \rwlx\}\\
\ExecConfs  &::= \Regs \times \Heaps \\
\Globals & ::= \{\glob, \local \} \\
\Caps  &::= (\Perms \times \Globals) \times \Addrs \times (\Addrs + \{ \infty \}) \times \Addrs\\
\Confs &::= \ExecConfs + \{\failed \} + \halted \times \ExecConfs \\
\HeapSegments &::= \Addrs \parfun \Words
\end{align*}
Local capabilities have been added by adding a new domain $\Globals$ which represents whether a capability is local or global. There are two new permissions $\readwritel$ and $\rwlx$ that permits writing local capabilities. They are otherwise the same as their non-''permit write local'' counterparts.

\begin{figure}[!h]
  \centering
  \begin{tikzpicture}[main node/.style={}]
  \node[main node] (1) {$\glob$};
  \node[main node] (2) [below of=1] {$\local$};

@@

\begin{table}[!h]
  \centering
  \begin{tabular}[!h]{r |  p{7cm} }
  $\noperm$ & No permissions. Grants no permissions\\
\hline
  $\readonly$ & Read only. Grants read permission \\
\hline
  $\readwrite$ & Read-write. Grants read and write permission. Storage of local capabilities prohibited. \\
\hline
  $\readwritel$ & Read-write, permit write local. Grants read and write permission. Storage of local capabilities possible. \\
\hline
  $\exec$ & Execute permission. Grants execute and read permissions.\\
\hline
  $\entry$ & Enter permission. This permission grants no access, but when jumped to, it will turn into an $\exec$ permission.\\
\hline
  $\rwx$ & Read-write-execute permission. Grants read, write, and execute permissions. Storage of local capabilities prohibited. \\
\hline
  $\rwlx$ & Read-write-execute, permit write local. Grants read, write, and execute permissions. Storage of local capabilities possible.
\end{tabular}

@@

\caption{Permission hierarchy}
\label{fig:perm-hier}
\end{figure}
Notation:
\[
\begin{array}{rcl}
i       &\in& \Instrs \\
r       &\in& \RegName\\
\pc     &\in& \Caps \\
\pcreg  &\in& \RegName \\
\Phi    &\in& \ExecConfs \\
m, \memheap&\in& \Heaps \\
\memreg &\in& \Regs \\
\addr   &\in& \Addrs\\
\perm   &\in& \Perms\\
((\perm,\gl),\start,\addrend,\addr) &\in& \Caps \\
n       &\in& \ints\\
\ms     &\in& \MemSegments
\end{array}
\]
Words and instructions:
\[
\begin{array}{rcl}
\lv    &::=& \refreg{r} \\
\hv    &::=& \refheap{r}\\
\rv    &::=& n \mid \lv \\
i      &::=& 
             \jmp{\lv} \mid 
             \jnz{\lv}{\rv} \mid
             \move{\lv}{\rv} \mid 
             \load{\lv}{\hv} \mid 
             \store{\hv}{\rv} \mid  \\
       &   & \plus{\lv}{\rv}{\rv} \mid 
             \lea{\lv}{\rv} \mid 
             \restricttwo{\lv}{\rv} \mid 
             \subseg{\lv}{\rv}{\rv} \mid  \\
       &   & \isptr{\lv}{\rv} \mid 
             \getp{\lv}{\lv} \mid 
             \getl{\lv}{\lv} \mid 
             \getb{\lv}{\lv} \mid
             \gete{\lv}{\lv} \mid
             \geta{\lv}{\lv} \mid \\
       &   & \fail \mid
             \halt \mid 
\end{array}
\]
Further define $\reg_0 \in \Regs$ such that
\[
  \forall r \in \RegName \ldotp \reg_0(r) = 0
\]

\subsection*{Semantics}
Assume a $\decode$ function that decodes words to instructions:
\begin{align*}
\decode &:\Words \fun \Instrs
\end{align*}
Assume an $\encodePerm$, $\encodeLoc$, and $\encodePermPair$ function that encodes a permissions, locality, and permission pair, respectively, as an integer:
\begin{align*}
\encodePerm &: \Perms \fun \ints \\
\encodeLoc &: \Globals \fun \ints \\
\encodePermPair &: (\Perms \times \Globals) \fun \ints \\
\end{align*}
Further, assume an inverse function, $\decodePermPair{}$, that decodes permissions
\[
  \decodePermPair{} : \ints \fun (\Perms \times \Globals)
\]
\begin{align*}
  \Phi & \rightarrow \sem{\decode(\memheap(\addr))}(\Phi) & &                                   
                                                              \arraycolsep=0pt
                                                              \begin{array}{l}
                                                                \text{if $\memreg(\pcreg) = \stdcap$}\\
                                                                \quad\text{and $\start \leq \addr \leq \addrend$}\\
                                                                \quad\text{and $\perm \in \{ \exec,\rwx, \rwlx \}$ }
                                                              \end{array}\\
\Phi & \rightarrow \failed                                 & & \text{otherwise}
\end{align*}

@@

                                                   \quad\text{and $\decodePermPair{n}\sqsubseteq \permp$}\\
                                                   \quad\text{and $c = (\decodePermPair{n},\start,\addrend,\addr)$}

@@

\lau{Do we need other arithmetic operators than plus?}

@@

                                                \text{if $\memreg(r) = \stdcap[\permp]$} \\

@@

                                                \quad\text{and $c = (\permp,n_1,n_2,\addr)$}

@@

                              \text{if $\memreg(r_2) = ((\_,\_),\_,\_,\addr)$}

@@

                              \text{if $\memreg(r_2) = ((\_,\_),\start,\_,\_)$}

@@

                              \text{if $\memreg(r_2) = ((\_,\_),\_,\addrend,\_)$}

@@

                              \text{if $\memreg(r_2) = ((\perm,\_),\_,\_,\_)$}

@@

                              \text{if $\memreg(r_2) = ((\_,\gl),\_,\_,\_)$}

@@

                            \stdUpdatePc{\updateReg{r_1}{1}} & \text{if $\memreg(\rv) \in \Caps$ } \\

@@

After adding local capabilities, most of the definitions have changed slightly to take into account the new form the capabilities take. The only two instructions that have undergone enough change to deserve to be mentioned are $\mathtt{store}$ and $\mathtt{restrict}$.

$\mathtt{store}$ has the same conditions as before, but if what we try to write is a local capability, then it needs to be a write access that permits write local.

$\mathtt{restrict}$ looks like before, but now it uses a ``permission pair'' which is the access permission and the locality. Whether a $\mathtt{restrict}$ succeeds depends on the ordering, so the real change lies in the new ordering, which is described further above (it is the point-wise ordering of the pair).

Define the following macros: $\mathtt{restrict}$, $\mathtt{subseg}$, and $\mathtt{lea}$ that does not overwrite the source register. A $\mathtt{store}$ that allows integers to be stored directly.
\begin{align*}
  \restrict{r_1}{r_2}{r_3} \; r_4 &\defeq
                                    \begin{aligned}[t]
                                      & \move{r_1}{r_2} \\
                                      & \restrict{r_1}{r_3}{r_4}
                                    \end{aligned} \\
  \subseg{r_1}{r_2}{r_3} \; r_4   &\defeq
                                    \begin{aligned}[t]
                                      & \move{r_1}{r_2} \\
                                      & \subseg{r_1}{r_3}{r_4}
                                    \end{aligned}\\
  \lea{r_1}{r_2} \; r_3           &\defeq
                                    \begin{aligned}[t]
                                      & \move{r_1}{r_2} \\
                                      & \lea{r_1}{r_3}
                                    \end{aligned}\\
  \store{r}{n} &\defeq
                   \begin{aligned}[t]
                     & \move{r_i}{n} \qquad \text{ where $r_i$ unused register}\\
                     & \store{r}{r_i}
                   \end{aligned}
\end{align*}

@@

\begin{specification}[Malloc v.2]
  $c_\malloc$ satisfies the specification for malloc iff
\[  
  \begin{aligned}
    &\exists \iota_{\malloc,0} \ldotp \\
    &\quad (\forall \iota' \futurestr \iota_{\malloc,0} \ldotp \forall W,i \ldotp W(i)=\iota' \Rightarrow \iota'.H (\iota'.s) (W) = \iota'.H (\iota'.s ([i \mapsto W(i)]))) \; \land \\
    &\quad \iota_{\malloc,0}.v = \perma \; \land \\
    &\quad (\forall \Phi \in \ExecConfs \ldotp \forall \ms_{\var{footprint}}, \hsframe \in \HeapSegments \ldotp \\
    &\qquad \forall i, n, \size \in \nats \ldotp \forall b,e,a \in \Addrs\ldotp\forall p \in \Perms \ldotp \\
    &\qquad \quad \forall \iota_\malloc \futurestr \iota_{\malloc,0} \land \\
    &\qquad \quad \memheap = \ms_{\var{footprint}} \uplus \hsframe \land\\
    &\qquad \quad \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
    &\qquad \quad \memreg(r_1) = \size \land \\
    &\qquad \quad \size \geq 0 \land \\
    &\qquad \quad \memreg(r_0) = (p,b,e,a) \land \\
    &\qquad \quad \memreg(\pcreg) = c_\malloc \\ %and this needs to change with ABI.
    &\qquad \quad \Rightarrow \\
    &\qquad \qquad\exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
    &\qquad \qquad \quad \exists j \in \nats \ldotp \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
    &\qquad \qquad \qquad \Phi \step[j] \Phi' \land \\
    &\qquad \qquad \qquad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
    &\qquad \qquad \qquad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
    &\qquad \qquad \qquad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
    &\qquad \qquad \qquad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
    &\qquad \qquad \qquad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{(p,b,e,a)}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
    &\qquad \qquad \qquad \size - 1 = e'-b' )
  \end{aligned}
\]
\end{specification}
In the specification above $\iota_\malloc'$ is a future region of the initial region that governs malloc.

\section{Macros}
In order to write readable example programs, we provide macros that can be programmed using the instruction set given in the formalisation.
\dominique{Similarly: how is malloc invoked?  Do we trust malloc enough to not encapsulate ourselves from it, i.e. provide an rx return capability and use callee-save registers?}
\lau{ I think this is a conceptual question as we can make it work with either. We already trust malloc to give us a fresh piece of memory and not reuse it later on, so we already assume malloc to be somewhat trusted, so why not go all the way? }
\dominique{what does ``fetch the capability ...'' mean?}
\lau{ We don't know where the capability resides, but if it is in memory, then it will be loaded into a register. }
\dominique{Wouldn't it be more clear to provide call with two explicit lists of registers: those which need to be stored, and those which are provided as arguments (i.e. which do not need to be erased)?}
\dominique{Perhaps you could also provide an explicit syntax for ``undefined symbols'' that should be filled in by a linker?}

\subsection{Linking and ABI}
In order to make capabilities to trusted code (and possibly untrusted code) available, we assume that some sort of linker has made these available. This is done in the following way: For every function, the first memory cell the capability for that function governs contains a capability for the linking table. Each function name in a program corresponds to an offset in the table, e.g., \texttt{malloc} could be at offset 0. When a name is used in a program, it indicates what entry from the linking table to pick. The table should always be accessible by taking a copy of the capability in the $\pcreg$-register and adjusting it to point to the first cell it governs.

The capability linking table can be shared between multiple functions that are linked to the same capabiblities as it is accessed through read-only capabilities.

\subsection{Flag table}
A function may use flags to signal failure. We use the convention that a flag table is available in the second memory cell of a functions code (so just after the linking table). The flag table is accessed through a read-write capability and initially it contains all zero. Like the linking table, each entry is associated with a name which may appear in the macros.

The flag table should never be shared between distrusting parties.

\subsection{Macro definitions}
\begin{description}
\item[\texttt{call} $f(\bar{r}_{\var{args}},\bar{r}_{\var{priv}})$] \forcenewline
$\bar{r}_{\var{args}}$ and $\bar{r}_{\var{priv}}$ are lists of registers. An overview of this call:
  \begin{itemize}
  \item Set up activation record
  \item Create local enter capability for activation (protected return pointer)
  \item Clear unused registers
  \item Jump
  \item Upon return: Run activation code
  \end{itemize}
A more detailed description of each of the above steps:
\begin{description}
  \item [Set up activation record]\forcenewline
    \begin{itemize}
    \item Run malloc to get a piece of memory with space for:
    \begin{itemize}
      \item Words in $\bar{r}_{\var{priv}}$
      \item Code return capability (opc)
      \item Activation code
    \end{itemize}
    \item Store the words in $\bar{r}_{\var{priv}}$ to the activation record. 
    \item Adjust a copy of the current pc to point to the return address in code and save it to the activation record.
    \item Write the activation code to the activation record.
  \end{itemize}
  \item [Create local enter capability for activation] Adjust the capability for the activation record to point to the beginning of the activation record and restrict it to a local enter-capability. Place this capability in $r_0$.
  \item [Clear unused registers]
    Clear all the register that are not $\pcreg$, $r_0$ or in $\bar{r}_{\var{args}}$.
  \item [Jump] Load the entry $f$ from the linking table to a free register and jump to it.
  \item [Activation code] The activation code does the following:
    \begin{itemize}
    \item Move the stored ``private'' words in to their respective $\bar{r}_{\var{priv}}$ registers.
%    \item Clear all registers but $r_1$, $\pcreg$, and $\bar{r}_{\var{priv}}$.
    \item Load the return capability to $\pcreg$
    \end{itemize}
\end{description}
\item[\texttt{malloc $r$ $n$}] Calls malloc to allocates a piece of memory of size $n$. The capability will be stored in register $r$.

\item[\texttt{assert$_{\var{flag}}$ $r_1$ $r_2$}] Compares the words in register $r_1$ and $r_2$ (if one of them is an integer, then use that in the comparison). If they are equal, then execution continues. If they are unequal, then the assertion flag named $\var{flag}$ in the flag list is set to 1 and execution halts (if no flag is specified, then the first flag in the list is set to 1).
\item[\texttt{cclear $r$}] Stores 0 to all the memory cells the capability $r$ governs.\footnote{This may in some cases seem like an unreasonable slow instruction. In a real system it would probably be implemented as a vector operation which allows modification of continuous segments of memory rather fast.}
\item[\texttt{rclear $\bar{r}$}] Moves 0 to all the registers in the list $\bar{r}$.
\end{description}
Note:
\begin{itemize}
\item \texttt{call} will fail if we have local capabilities in one of the registers of the ``private'' register list as it relies on a capability returned by malloc which will not be permit-write-local. Below, we introduce \texttt{scall} which can handle local capabilities in the ``private'' state.
\end{itemize}
\subsection{Stack}
Some programs will assume access to a stack which will be in part indicated by the program macros but also in the correctness lemma. The stack is accessed through a local $\rwlx$-capability. Programs will assume that the stack resides in some register, say $r_{\var{stk}}$.

The stack resides entirely in memory. There is no separation between the memory and the stack, so when we talk about the stack it is as a conceptual thing.
